%
%	Bibliography secion
%	TODO: bibtex
%

\begin{thebibliography}{9}

	\bibitem{map-reduce} J. Dean, S. Ghemawat. 
		\emph{MapReduce: Simplified Data Processing on Large Clusters}, \\
		OSDI '04, p. 137--150.
		\samehref{http://www.usenix.org/events/osdi04/tech/dean.html}

	\bibitem{cloud-haskell} J. Epstein, A.P. Black, S.P. Jones.
		\emph{Towards Haskell in the Cloud},\\
		\samehref{http://research.microsoft.com/en-us/um/people/simonpj/papers/parallel/remote.pdf}
		
	\bibitem{HdpH} P. Maier and P. Trinder.
		\emph{Implementing a High-level Distributed-Memory Parallel Haskell in Haskell},\\
		\samehref{http://www.macs.hw.ac.uk/~trinder/papers/HdpH.pdf}
		
	\bibitem{par-monad} S. Marlow, R. Newton, S.P. Jones.
		\emph{A monad for deterministic parallelism},\\
		\samehref{http://research.microsoft.com/en-us/um/people/simonpj/papers/parallel/monad-par.pdf}
		
	\bibitem{ShmStreaming} L. Lai, J. Zhou, L. Zheng, H. Li, Y. Lu, F. Tang, M. Guo.
		\emph{ShmStreaming: A Shared Memory Approach for Improving Hadoop Streaming Performance},\\
		\samehref{http://www.computer.org/csdl/proceedings/aina/2013/4953/00/4953a137-abs.html}
			
	\bibitem{HaLoop} Y. Bu, B. Howe, M. Balazinska, M. D. Ernst.
		\emph{HaLoop: Efficient Iterative Data Processing on Large Clusters},\\
		\samehref{http://dl.acm.org/citation.cfm?id=1920881}
		
	\bibitem{IncMR} C. Yan, X. Yang, Z. Yu, M. Li, X. Li.
		\emph{IncMR: Incremental Data Processing based on MapReduce},\\
		\samehref{http://www.bibsonomy.org/bibtex/29cb17af55bcdf7a99bfd4e0e62bccf73/dblp}
	
	\bibitem{fsharp-compexpr} T. Petricek, D. Syme.
		\emph{Syntax Matters: Writing abstract computations in \fsharp},\\
		Pre-proceedings of TFP, 2012.\\
		\href{http://www.cl.cam.ac.uk/~tp322/drafts/notations.pdf}
			{\texttt{http://www.cl.cam.ac.uk/{\textasciitilde}tp322/drafts/notations.pdf}}

	\bibitem{fsharp-async} D. Syme, T. Petricek T, D. Lomov.
		\emph{The \fsharp{} Asynchronous Programming Model}. \\ 
		\samehref{http://research.microsoft.com/apps/pubs/default.aspx?id=147194}.
		
	\bibitem{data-types-ala-carte} 
		W. Swierstra. \emph{Data types \`{a} la carte}, Journal of Functional Programming, \\ 	   
		Cambridge University Press, 2008. \\
		\href{http://www.cs.ru.nl/~W.Swierstra/Publications/DataTypesALaCarte.pdf}
			{\texttt{http://www.cs.ru.nl/{\textasciitilde}W.Swierstra/%
						Publications/DataTypesALaCarte.pdf}}
		
	\bibitem{scala-trampolines}
		R.O. Bjarnarson. \emph{Stackless Scala With Free Monads}. Scala Days 2012.\\
		\samehref{http://blog.higher-order.com/assets/trampolines.pdf}

\end{thebibliography}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "mbrace"
%%% End: 
