%
%	Appendix
%

\section*{Appendix}

\subsection*{Future Work}

\Mbrace{} is a large project that faces multiple technical challenges. As such,
it remains (as of the private alpha release) a work in progress with many important
features still under development. In this section we discuss some of the features
which we consider to be important milestones in the future of the \mbrace{} project.

\subsubsection*{\csharp{}-LINQ support}
CloudLinq% 
\footnote{\samehref{https://github.com/nessos/CloudLinq}}
is a satellite project for \mbrace{} aiming at providing an idiomatic 
API for the immensely popular \csharp{} language in the
form of a LINQ provider%
\footnote{\samehref{http://msdn.microsoft.com/en-us/library/vstudio/bb397926.aspx}}
for data parallelism. This would allow building cloud workflows using simple
LINQ-style queries:
\begin{lstlisting}[language=CSharp]
var result =  from c in Customers.AsCloudQueryExpr()
			  where c.City == "Athens"
			  orderby c.Name
			  select c.Name, c.City
\end{lstlisting}
CloudLinq can achieve excellent performance by taking advantage of LinqOptimizer%
\footnote{\samehref{https://github.com/nessos/LinqOptimizer}}
which is an optimizer for Linq and compiles declarative queries into fast imperative code. Another interesting direction is the combination
of CloudLinq with GpuLinq%
\footnote{\samehref{https://github.com/nessos/GpuLinq}}
in order to compile queries to OpenCL kernel code for fast execution on GPU powered cloud instances.



\subsubsection*{Mono Support}

The Mono project%
\footnote{The Mono project, \samehref{http://www.mono-project.com/}.}
is a popular open source implementation of Microsoft's \dotnet{} framework.
The Mono framework is in the unique position of offering a truly cross-platform
\dotnet{} development experience as it supports most major operating systems,
including Linux and Mac OS X.

Just like the \fsharp{} language itself, we believe that a big part of \mbrace{}'s 
future lies with Mono and the open source community in general. Providing support
for the Mono framework is strategically important since it opens up the market
to diverse developer cultures and data center infrastructures.


% TODO: fill out the following sections

%\subsection*{Web UI}

%\subsection*{Internals Discussion}
%
%\subsubsection*{The Distributed Actor Framework}
%
%\subsubsection*{Serialization}
%
%\subsubsection*{The Monadic Trampoline}



